### 2018-07-18 paper.tex
Initial Commit - outline and references

\documentclass[psamsfonts]{amsart}

% -------Packages---------
\usepackage{microtype}[final]
\usepackage[style=numeric]{biblatex}
\usepackage{hyperref}
\addbibresource{./bib.bib}
\usepackage{amssymb,amsfonts}
\usepackage[all,arc]{xy}
\usepackage{enumerate}
\usepackage{mathrsfs}

%--------Theorem Environments--------
%theoremstyle{plain} --- default
\newtheorem{thm}{Theorem}[section]
\newtheorem{cor}[thm]{Corollary}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{conj}[thm]{Conjecture}
\newtheorem{quest}[thm]{Question}
\newtheorem{prob}[thm]{Problem}

\theoremstyle{definition}
\newtheorem{defn}[thm]{Definition}
\newtheorem{defns}[thm]{Definitions}
\newtheorem{con}[thm]{Construction}
\newtheorem{exmp}[thm]{Example}
\newtheorem{exmps}[thm]{Examples}
\newtheorem{notn}[thm]{Notation}
\newtheorem{notns}[thm]{Notations}
\newtheorem{addm}[thm]{Addendum}
\newtheorem{exer}[thm]{Exercise}

\theoremstyle{remark}
\newtheorem{rem}[thm]{Remark}
\newtheorem{rems}[thm]{Remarks}
\newtheorem{warn}[thm]{Warning}
\newtheorem{sch}[thm]{Scholium}

\makeatletter
\let\c@equation\c@thm
\makeatother
\numberwithin{equation}{section}

\title{Undecidability and the structure of the Turing degrees}

\author{Philip Adams}

\date{\today}

\begin{document}

\begin{abstract}

  This paper explores the structure and properties of the Turing degrees, or
  degrees of undecidability. We introduce the Turing machine, an abstract model
  of computation, in order to develop the concepts of undecidability and Turing
  reduction. We demonstrate the technique of proof by reduction through a series
  of examples of undecidable problems related to context-free grammars. We then
  employ reducibility to consider a partial ordering on the set of Turing degrees, $\mathcal{D}$. Finally, we prove a variety of theorems
  related to the structure of $\mathcal{D}$.

\end{abstract}

\maketitle

\tableofcontents

\section{Introduction}
Historical overview of the field, motivation for study
\cite{ambos-spies06:_degrees_unsol}
\cite{soare16_turin_comput}
\cite{soare1999history}
\cite{lerman16:_degrees_unsol}

\section{Turing Machines}
% Turing 1936?
\cite{turing37_comput_number_with_applic_to_entsc}
\section{Undecidability}

\begin{prob}[The Halting Problem]
  \cite{turing37_comput_number_with_applic_to_entsc}
  \cite{sipser13:_introd_theor_comput}
\end{prob}
\begin{thm}[Rice's Theorem] \cite{kozen99_autom}
  % Proof by reduction to the halting problem?
\end{thm}
\subsection{Incompleteness}
\begin{thm}[G\"odel's Incompleteness Theorem] 
  % Kleene 1943? Also shown in hopcroft 1979
  \cite{kleene43_recur_predic_quant}
\end{thm}
\section{Reducibility}
\cite{sipser13:_introd_theor_comput}
\cite{post44:_recur}

\cite{kleene80_introd}

\section{Context Free Grammars}

\begin{defn}
 A Context Free Grammar is...
\end{defn}

\begin{defn}
  A Pushdown Automata is...
\end{defn}

\begin{thm}[Equivalence of CFGs and Pushdown Automata]
  \cite{hopcroft07:_introd_autom_theor_languag_comput}
\end{thm}
\begin{thm}[Closure Properties]
  \cite{sipser13:_introd_theor_comput}
  Union, Concat, Star, intersection with regular, substitution
\end{thm}
\begin{prob}
Empty, finite
\end{prob}
\subsection{Undecidable Problems}
\begin{prob}[The Post Correspondence Problem]
\cite{hopcroft07:_introd_autom_theor_languag_comput}
\end{prob}
\begin{prob}
 Undecidability of ambiguity 
\end{prob}
\cite{greibach66:_unsol_recog_linear_contex_free_languag}
\cite{Hopcroft1969}
\begin{prob}
  Universality
\end{prob}
\begin{prob}
  Equality, inclusion
\end{prob}
\begin{prob}
  Disjointness
\end{prob}
\subsubsection{Problems decidable for Deterministic CFGs}
\cite{ginsburg65:_deter}
\begin{defn}
  A Deterministic Context Free Grammar is...
  \cite{sipser13:_introd_theor_comput}
\end{defn}

\begin{thm}[Closure Properties]
  \cite{sipser13:_introd_theor_comput}
  Complement
\end{thm}

\begin{prob}[The Equivalence Problem for Deterministic CFGs]
\cite{senizergues_det_pd_decid}
\end{prob}

\begin{prob}
  Universality
\end{prob}

\section{Turing Degrees}
\begin{defn}
  A Turing degree...
\cite{post44:_recur}
\cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
\end{defn}

\begin{defn}
  The jump operator...

\end{defn}

\begin{defn}
  Computably Enumerable
\end{defn}

\begin{defn}
  Completeness
\end{defn}

\subsection{Properties and Structure}
\begin{prob}[Post's Problem]
  \cite{post44:_recur}
  % Marchenkov 1976
  % Independently by Friedberg 57, Mucnik 56/58 using priority method
  % Priority free solution Kucer 1986''
  % ``Natural'' solns don't exist
  \cite{Friedberg236}
\end{prob}

% Myhill 56 c.e.lattice

\begin{lem}
  D does not form a lattice
  \cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
\end{lem}
\begin{thm}
  D forms an upper semi-lattice
  % Kleene-Post 54 Degrees upper semi-lattice, jump an operator on degrees
  \cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
\end{thm}

\begin{thm}
  Existence of minimal degrees
  % Spector 56,Shoenfield 66 (Better)
  \cite{spector56_degrees_recur_unsol}
  \cite{shoenfield66_theor_minim_degrees}
\end{thm}

\begin{thm}
  The c.e. degrees are dense
  \cite{sacks64:_recur_enumer_degrees_dense}
\end{thm}

Homogeneity problems
% Feiner 1970, Shore 1979

Theory of D Undecidable

\subsection{Turing Degrees of Problems related to CFGs}
\cite{REEDY197577}

\section*{Acknowledgments}
It is a pleasure to thank my mentor, Ronno Das, for supervising this project and
providing valuable feedback and advice. I would also like to thank Peter May for
organizing this REU, and Daniil Rudenko for running the Apprentice Program.

\printbibliography

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2018-08-09 paper.tex
Fixing bibliography
Readings complete, starting work on the CFG section

@@
 \usepackage{hyperref}
-\addbibresource{./bib.bib}
+\addbibresource{/home/philip/documents/text/references/bibliography.bib}
 \usepackage{amssymb,amsfonts}
@@
 \usepackage{mathrsfs}
+\usepackage{todonotes}
 
@@
 \cite{lerman16:_degrees_unsol}
+\subsection{Decision Problems}
 
@@
 \begin{prob}[The Halting Problem]
+  \label{prob:halting}
   \cite{turing37_comput_number_with_applic_to_entsc}
@@
 \end{thm}
-\section{Reducibility}
+\subsection{Reducibility}
+While it is possible to prove that a problem is undecidable directly, as in
+problem~\ref{prob:halting}, it is often more convenient to prove undecidability
+through comparison to problems which are already known to be undecidable. This
+comparison takes place through the technique of Turing reduction.
+\begin{defn}
+  Given two decision problems $A$ and $B$, we say that $A$ is Turing reducible
+  to $B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
+  a machine $D_A$ that decides $A$.
+\end{defn}
+So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
+because any solution to $B$ also leads to a solution of $B$. So, if we have some
+problem $B$ that we would like to prove is undecidable, we can do so by showing
+that some problem $A$ which is already known to be undecidable is reducible to
+$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
+undecidable. Alternatively, if we would like to show that some problem $A$ is
+decidable, it is sufficient to show that it is reducible to some problem $B$
+that is known to be decidable.
 \cite{sipser13:_introd_theor_comput}
@@
 \begin{prob}
-Empty, finite
+Empty, finite, containment within regular
 \end{prob}
@@
 \begin{prob}[The Post Correspondence Problem]
-\cite{hopcroft07:_introd_autom_theor_languag_comput}
+  Consider some alphabet $\Sigma$ and two finite lists of words over $\Sigma$
+  denoted $A = a_1,\dots,a_n$ and $B = b_1, \dots, b_n$. Then, does there exist
+some sequence of indices $(i_k),$ for $1 \leq k \leq K$ and for
+  $K\geq 1$, and with $1 \leq i_k \leq n$ for all $k$ such that
+  \[
+    a_{i_1}\dots a_{i_K} = b_{i_1}\dots b_{i_K}?
+  \]
+  \begin{proof}[Solution]
+    This problem is shown to be undecidable through a reduction to the halting
+    problem. The proof involves \todo{Describe proof in detail}
+    \cite{sipser13:_introd_theor_comput}
+  \end{proof}
 \end{prob}
-\begin{prob}
- Undecidability of ambiguity 
+
+The Post correspondence problem is a useful tool, because it allows us to
+demonstrate the undecidability of problems without doing the complex reasoning
+about Turing machines that the halting problem requires. We will now show the
+undecidability of a variety of problems about context-free grammars through
+reduction to the Post correspondence problem.
+
+\begin{prob}[Disjointness]
+  Given two context-free grammars $P,Q$, is $L(P)\cap L(Q) = \varnothing$?
+  \begin{proof}[Proof of Undecidability]
+    Consider the Post correspondence problem for two lists of words $A,B$. We
+    construct two context free grammars from these lists as follows:
+    \begin{equation*}
+      \begin{split}
+        G_A \rightarrow  &\;a_11 \\
+        &\!\!\!\!\vdots \\
+        G_A \rightarrow  &\;a_nn \\
+        G_A \rightarrow  &\;a_1G_A1 \\
+        &\!\!\!\!\vdots \\
+        G_A \rightarrow  &\;a_nG_An \\
+      \end{split}
+      \qquad
+      \begin{split}
+        G_B \rightarrow  &\;a_11 \\
+        &\!\!\!\!\vdots \\
+        G_B \rightarrow  &\;a_nn \\
+        G_B \rightarrow  &\;a_1G_B1 \\
+        &\!\!\!\!\vdots \\
+        G_B \rightarrow  &\;a_nG_Bn \\
+      \end{split}
+    \end{equation*}
+    Then, we can observe that for some string $s$ to exist in both $L(G_A)$ and $L(G_B)$,
+    it must be a solution to the Post correspondence for $A,B$. So, if $L(G_A)\cap
+    L(G_B)=\varnothing$, then there are no solutions to the Post correspondence
+    for $A,B$. So, it follows that the Post correspondence problem is reducible
+    to the problem of the disjointness of context-free grammars, so since the
+    Post correspondence problem is undecidable, it follows that the disjointness
+    problem is undecidable.
+  \end{proof}
+\end{prob}
+
+\begin{prob}[Universality]
+  \label{prob:cfg:universality}
+  Take some context-free grammar $G$ over an alphabet $\Sigma$. Then, is
+  \[
+    L(G) = \Sigma^*?
+  \]
+  \begin{proof}[Proof of Undecidability]
+
+  \end{proof}
+\end{prob}
+
+\begin{defn}
+An ambiguous grammar is...
+\end{defn}
+
+\begin{prob}[Ambiguity]
+  Take some context free grammar $G$. Is $G$ ambiguous?
+  \begin{proof}[Proof of Undecidability]
+    Consider the Post correspondence problem for two lists of words
+    $A,B$. Construct their corresponding grammars $G_A,G_B$. Now, consider the
+    grammar
+    \[
+      G \rightarrow G_A \vert G_B.
+    \]
+    It follows that the ambiguity of $G$ implies that a solution to the Post
+    correspondence problem for $A,B$ exists, so the Post correspondence problem
+    reduces to the ambiguity problem, so the ambiguity problem is undecidable.
+  \end{proof}
 \end{prob}
@@
 \begin{prob}
-  Universality
+  Regularity
 \end{prob}
-\begin{prob}
-  Equality, inclusion
+\begin{prob}[Equality]
+  Take context free grammars $G_1,G_2$. Is $L(G_1)=L(G_2)$?
+  \begin{proof}[Proof of Undecidability]
+    Note that $\Sigma^*$ is regular, so it is also a context-free. So,
+    we have that Problem~\ref{prob:cfg:universality} reduces to the equality
+    problem. It follows that the equality problem is undecidable.
+  \end{proof}
 \end{prob}
-\begin{prob}
-  Disjointness
+\begin{prob}[Inclusion]
+  Take context free grammars $G_1,G_2$. Is $L(G_1)\subseteq L(G_2)$?
+  \begin{proof}[Proof of Undecidability]
+    As before, note that $\Sigma^*$ is regular, so it is also a
+    context-free. Additionally, observe that for any grammar $G$, $L(G)\subseteq
+    \Sigma^*$. So,
+    we have that Problem~\ref{prob:cfg:universality} reduces to the inclusion
+    problem, since $\Sigma^* \subseteq L(G) \implies \Sigma^* = L(G)$. It follows that the inclusion problem is undecidable.
+  \end{proof}
 \end{prob}
+
 \subsubsection{Problems decidable for Deterministic CFGs}

### 2018-08-09 paper.tex
regular containment problem

@@
 \end{thm}
-\begin{prob}
-Empty, finite, containment within regular
+\begin{prob}[Emptiness]
+  \label{prob:cfg:emptiness}
+Take some context-free grammar $G$. Is $L(G)=\varnothing$?
 \end{prob}
+
+\begin{prob}[Finite]
+Take some context-free grammar $G$. Is $L(G)$ finite?
+\end{prob}
+
+\begin{prob}[Regular Containment]
+Take some context-free grammar $G$ and some regular language $R$. Is
+$L(G)\subseteq R$?
+\begin{proof}[Proof of Decidability]
+  We have that $L(G)\subseteq R$ if and only if
+  $\overline{L(G)}\cup R = \Sigma^*$. So, it follows that $L(G)$ is contained by
+  $R$ if and only iff $\overline{\overline{L(G)}\cup R} = \varnothing$. By
+  DeMorgan's laws, we have that this is equivalent to the statement
+  $L(G)\cap \overline{R} = \varnothing$. Regular languages are closed under
+  complement and context-free languages are closed under intersection with a
+  regular language, so it follows that $L(G)\cap \overline{R}$ is a context free
+  language. So the problem of containment by a regular language is reducible to
+  Problem~\ref{prob:cfg:emptiness}, the problem of emptiness, which is
+  decidable. So, the problem of containment by a regular language is decidable.
+\end{proof}
+
+\end{prob}
+
 \subsection{Undecidable Problems}
@@
 \begin{prob}[Ambiguity]
-  Take some context free grammar $G$. Is $G$ ambiguous?
+  Take some context-free grammar $G$. Is $G$ ambiguous?
   \begin{proof}[Proof of Undecidability]
@@
 \begin{prob}
-  Regularity
+  Take some context-free grammar $G$. Is $L(G)$ regular?
+  \begin{proof}[Proof of Undecidability]
+    Note that $\Sigma^*$ is regular, so if $L(G)$ is not regular then it follows
+    that $L(G)\neq \Sigma^*$. Additionally, universality is decidable within
+    regular languages, because they are closed under complement and emptiness is
+    decidable even within context-free grammars. So, we have that
+    Problem~\ref{prob:cfg:universality} reduces to the regularity problem. It
+    follows that the regularity problem is undecidable.
+  \end{proof}
 \end{prob}

### 2018-08-10 paper.tex
Discussion of FSAs, PDAs

@@
 \usepackage{todonotes}
+\usepackage{tikz}
+\usetikzlibrary{automata,positioning}
+\usepackage[nolist]{acronym}
 
@@
 \tableofcontents
+\begin{acronym}
+  \acro{PDA}[PDA]{Pushdown Automaton}
+  \acroplural{PDA}[PDAs]{Pushdown Automata}
+  \acro{DPDA}[DPDA]{Deterministic Pushdown Automaton}
+  \acroplural{DPDA}[DPDAs]{Deterministic Pushdown Automata}
+  \acro{TM}[TM]{Turing Machine}
+  \acro{FSA}[FSA]{Finite State Automaton}
+  \acroplural{FSA}[FSAs]{Finite State Automata}
+  \acro{CFG}[CFG]{Context-Free Grammar}
+  \acro{CFL}[CFL]{Context-Free Language}
+\end{acronym}
 
@@
 \section{Turing Machines}
-% Turing 1936?
-\cite{turing37_comput_number_with_applic_to_entsc}
+Upon consideration of different types of problems, it quickly becomes clear that
+some are much more difficult than others. For example, when considering words
+over the alphabet $\{a,b,c\}$, it seems much easier to determine whether a word
+ends in an $a$ than whether a word is a palindrome. It seems even harder still
+to determine whether the word is of the form $a^nb^nc^n$. In order to formalize
+these notions of difficulty, we need to build abstract models of computation,
+and then test their ability to decide such questions. We observe that a machine
+that can solve the first problem only needs to ``remember'' the same amount of
+information no matter how long the input is: the last letter of the word. In
+contrast, the amount of memory that the second and third problems require is
+dependent on the size of the input, because the first half of the palindrome or
+$n$, respectively, can be arbitrarily large. The difference between the second
+and third problems is more subtle: the second problem can be solved moving only
+one direction in memory, while the third problem requires the ability to ``look
+back'' in memory.
+\par
+Based on these differences, we can begin to construct different models of
+computation that have just enough power to solve each problem. The first problem
+can be solved by a \emph{\ac{FSA}}, a collection of a finite
+number of states, a transition function, and a set of accepting states. The
+\ac{FSA} accepts the input if the sequential application of each element to the
+input finishes in an accepting state, and rejects otherwise. \aclp{FSA} are often represented by charts as in Figure~\ref{fig:fsa}, a
+representation of an \ac{FSA} that decides the first problem.
+\begin{figure}[h]
+  \caption{An \ac{FSA} that determines whether a string ends in
+    `$a$'.}
+  \label{fig:fsa}
+  \begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
+  \node[state,initial] (q_0) {$q_0$};
+  \node[state,accepting] (q_1) [right=of q_0]{$q_1$};
+  \path[->]
+  (q_0) edge [loop above] node {b,c} ()
+  edge [bend left] node  {a} (q_1)
+  (q_1) edge [bend left] node {b,c} (q_0)
+  edge [loop above] node {a} ();
+\end{tikzpicture}
+\end{figure}
+\par
+The second problem can be solved by a machine called a \emph{\ac{PDA}}, which is essentially a finite state automaton with the addition of
+a stack. Rather than transforming the current state and the input into a new
+state, as with a finite state automata, the transition function of a \ac{PDA} also considers the stack, which it can pop and push from. Within the
+collection of \acp{PDA}, there is another division between
+\emph{\acp{DPDA}} and \emph{Nondeterministic \aclp{PDA}}. For \acp{DPDA}, the
+transition function only outputs one move, rather than a set of
+moves. Nondeterministic \acp{PDA} are able to recognize more languages than
+\acp{DPDA}. Figure~\ref{fig:pda} is a graphical representation of a \ac{PDA}
+that decides palindromes. In this representation, $\epsilon$ is an empty input,
+and  is a special symbol that denotes the bottom of the stack. The moves are
+represented by a double of input and stack operation.
+
+\begin{figure}[h]
+  \label{fig:pda}
+  \caption{A \acl{PDA} that decides palindromes over the alphabet ${a,b}$}
+  \begin{tikzpicture}[shorten >=1pt,node distance=3cm,on grid,auto]
+  \node[state,initial] (q_0) {$q_0$};
+  \node[state] (q_1) [right=of q_0]{$q_1$};
+  \node[state] (q_2) [right=of q_1]{$q_2$};
+  \node[state,accepting] (q_3) [right=of q_2]{$q_3$};
+  \path[->]
+  (q_0) edge node {$\epsilon,\epsilon\to\$$} (q_1)
+  (q_1) edge [loop above] node[text width=1.5cm,align=center] {$0,\epsilon\to 0$
+    $1,\epsilon\to 1$} ()
+  edge node {$\epsilon,\epsilon\to\epsilon$} (q_2)
+  (q_2) edge [loop above] node[text width=1.5cm,align=center] {$0,0\to\epsilon$
+    $1,1\to\epsilon$} ()
+  edge node {$\epsilon,\$\to\epsilon$} (q_3);  
+  \end{tikzpicture}
+\end{figure}
+\par
+Finally, we arrive at the \ac{TM}, devised by Alan Turing in his paper \emph{On
+  Computable Numbers, With an Application To the Entscheidungsproblem}
+~\cite{turing37_comput_number_with_applic_to_entsc}. \aclp{TM} offer an
+unrestricted model of computation,\todo{Discussion of TMs}
 \section{Undecidability}
@@
   \label{prob:cfg:emptiness}
-Take some context-free grammar $G$. Is $L(G)=\varnothing$?
+Let $G$ be a context-free grammar. Is $L(G)=\varnothing$?
 \end{prob}
@@
 \begin{prob}[Finite]
-Take some context-free grammar $G$. Is $L(G)$ finite?
+Let $G$ be a context-free grammar. Is $L(G)$ finite?
 \end{prob}
@@
     This problem is shown to be undecidable through a reduction to the halting
-    problem. The proof involves \todo{Describe proof in detail}
-    \cite{sipser13:_introd_theor_comput}
+    problem. The proof is technical, and involves encoding the computation
+    history of a Turing machine on an input in such a way that the encoding satisfies the
+    PCP if and only if the Turing machine accepts the input. It is described in
+    detail by Sipser in his book \emph{Introduction to the Theory of Computation}
+    \cite{sipser13:_introd_theor_comput}.
   \end{proof}
@@
 \begin{prob}[Disjointness]
-  Given two context-free grammars $P,Q$, is $L(P)\cap L(Q) = \varnothing$?
+  Let $P,Q$ be context-free grammars. Is $L(P)\cap L(Q) = \varnothing$?
   \begin{proof}[Proof of Undecidability]
@@
   \label{prob:cfg:universality}
-  Take some context-free grammar $G$ over an alphabet $\Sigma$. Then, is
+  Let $G$ be some context-free grammar over an alphabet $\Sigma$. Then, is
   \[
@@
 \begin{prob}[Ambiguity]
-  Take some context-free grammar $G$. Is $G$ ambiguous?
+  Let $G$ be a context-free grammar. Is $G$ ambiguous?
   \begin{proof}[Proof of Undecidability]
@@
 \begin{prob}
-  Take some context-free grammar $G$. Is $L(G)$ regular?
+  Let $G$ be a context free grammars. Is $L(G)$ regular?
   \begin{proof}[Proof of Undecidability]
@@
 \begin{prob}[Equality]
-  Take context free grammars $G_1,G_2$. Is $L(G_1)=L(G_2)$?
+  Let $G_1,G_2$ be context free grammars. Is $L(G_1)=L(G_2)$?
   \begin{proof}[Proof of Undecidability]
@@
 \begin{prob}[Inclusion]
-  Take context free grammars $G_1,G_2$. Is $L(G_1)\subseteq L(G_2)$?
+  Let $G_1,G_2$ be context free grammars. Is $L(G_1)\subseteq L(G_2)$?
   \begin{proof}[Proof of Undecidability]
@@
 
-\subsubsection{Problems decidable for Deterministic CFGs}
+\subsubsection{Problems decidable for Deterministic CFLs}
 \cite{ginsburg65:_deter}
 \begin{defn}
-  A Deterministic Context Free Grammar is...
-  \cite{sipser13:_introd_theor_comput}
+  A deterministic context-free language is a language accepted by a
+  deterministic pushdown automata.
 \end{defn}
@@
 \begin{thm}[Closure Properties]
-  \cite{sipser13:_introd_theor_comput}
-  Complement
+  If $L$ is a deterministic context-free language, then $\overline{L}$ is a
+  deterministic context-free language.
 \end{thm}
 
-\begin{prob}[The Equivalence Problem for Deterministic CFGs]
-\cite{senizergues_det_pd_decid}
+\begin{prob}
+  Let $L$ be a deterministic context-free language. Is $L = \Sigma^*$?
 \end{prob}
 
-\begin{prob}
-  Universality
+The decidability of the following problem was an open problem in the field of
+computability theory from 1965, when it was introduced by Ginsburg and Greibach \cite{ginsburg65:_deter}. In 1997, when it was solved by G\'eraud S\'enizergues
+\cite{senizergues_det_pd_decid}. A sketch of the proof follows.
+
+\begin{prob}[The Equivalence Problem for Deterministic CFGs]
+  Let $L_1,L_2$ be deterministic context-free languages. Is $L_1=L_2$?
+
 \end{prob}

### 2018-08-12 paper.tex
Expanding outline for Turing degree section

@@
 \newtheorem{exer}[thm]{Exercise}
+\newtheorem{fact}[thm]{Fact}
 
@@
 \subsection{Reducibility}
+\label{subsec:reducibility}
 While it is possible to prove that a problem is undecidable directly, as in
@@
 \begin{defn}
-  Given two decision problems $A$ and $B$, we say that $A$ is Turing reducible
-  to $B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
+  Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
+  to $B$ and write $A \leq_T B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
   a machine $D_A$ that decides $A$.
@@
   \]
-  \begin{proof}[Proof of Undecidability]
+  \begin{proof}[Proof of Undecidability]\todo{}
 
@@
 \section{Turing Degrees}
+In Subsection~\ref{subsec:reducibility}, we introduced the concept of
+Turing reducibility, denoted by the symbol $\leq_T$. This symbol suggests some type of
+ordering over the set of decision problems. In this section, we introduce that
+partial ordering on the set of decision problems and examine its properties.
+
 \begin{defn}
-  A Turing degree...
-\cite{post44:_recur}
-\cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
+  Let $A$ and $B$ be decision problems. We say that $A$ and $B$ are
+  \emph{mutually reducibile} and write $=_T$ if $A \leq_T B$ and $B \leq_T A$.
 \end{defn}
 
-\begin{defn}
-  The jump operator...
+\begin{fact}
+  The relation $=_T$ is an equivalence relation.
+\end{fact}
 
-\end{defn}
+We call the equivalence classes produced by this relation \emph{Turing degrees}
+or \emph{degrees of unsolvability}, a concept first introduced in
+1944 by Emil Post~\cite{post44:_recur}. We write the set of all such degrees as $\mathcal{D}$.
+
+\begin{lem}
+  The relation $\leq_T$ is a partial ordering of $\mathcal{D}$.
+\end{lem}
 
@@
 \subsection{Properties and Structure}
+\begin{lem}
+  The set $\mathcal{D}$ is of cardinality $2^{\aleph_0}$.
+\end{lem}
+
+\begin{lem}
+  Every Turing degree contains countably infinite elements.
+\end{lem}
+
+While it is simple to explicitly state problems lying in the lower degrees,
+in order to more easily study the general structure of $\mathcal{D}$, we
+introduce a new operator, \emph{jump}, that increments Turing degrees. Much of
+the behavior of this operator was shown in a join paper by Kleene and Post~\cite{kleene54_upper_semi_lattic_degrees_recur_unsol}.
+
+\begin{defn}
+  The jump operator
+\end{defn}
+
 \begin{prob}[Post's Problem]
@@
 
+Theory of D equivalent to to second-order arithmetic \cite{simpson77:_first_order_theor_degrees_recur_unsol}
+
 \subsection{Turing Degrees of Problems related to CFGs}

### 2018-08-13 paper.tex
Turing machine def

@@
 
-\section{Turing Machines}
+\section{Models of Computation}
 Upon consideration of different types of problems, it quickly becomes clear that
@@
   (q_0) edge node {$\epsilon,\epsilon\to\$$} (q_1)
-  (q_1) edge [loop above] node[text width=1.5cm,align=center] {$0,\epsilon\to 0$
-    $1,\epsilon\to 1$} ()
+  (q_1) edge [loop above] node[text width=1.5cm,align=center] {$a,\epsilon\to a$
+    $b,\epsilon\to b$} ()
   edge node {$\epsilon,\epsilon\to\epsilon$} (q_2)
-  (q_2) edge [loop above] node[text width=1.5cm,align=center] {$0,0\to\epsilon$
-    $1,1\to\epsilon$} ()
+  (q_2) edge [loop above] node[text width=1.5cm,align=center] {$a,a\to\epsilon$
+    $b,b\to\epsilon$} ()
   edge node {$\epsilon,\$\to\epsilon$} (q_3);  
@@
 \end{figure}
-\par
+
+\subsection{Turing Machines}
 Finally, we arrive at the \ac{TM}, devised by Alan Turing in his paper \emph{On
@@
 ~\cite{turing37_comput_number_with_applic_to_entsc}. \aclp{TM} offer an
-unrestricted model of computation,\todo{Discussion of TMs}
+unrestricted model of computation, equivalent to the general purpose computers
+that we are familiar with today. This equivalence is expressed by a pair of
+concepts called \emph{Turing completeness} and \emph{Turing equivalence}. A
+system of computation is Turing complete if it is able to simulate an arbitrary
+Turing machine, and is Turing Equivalent if there is a Turing machine that can
+simulate it. Essentially, Turing completeness is being \emph{at least} as
+powerful as Turing machines, while Turing equivalence is being \emph{exactly} as
+powerful as a Turing machine. Turing machines are themselves Turing complete, as are other
+abstract models of computation such as the \emph{lambda calculus}. General
+purpose programming languages are also considered Turing complete, because
+although they are run on real machines with finite storage space, this space is
+theoretically unbounded, and thus the expressive power of the languages are
+unbounded even if any particular instance of a program run on an actual machine
+would not be able to simulate all Turing machines. 
+\par
+A basic informal definition of a Turing machine is a collection of an infinite tape
+of discrete cells,
+a head that can move back and forth between cells on the tape as well as read and write
+from the tape, a register that stores the current state of the machine out of a
+finite set of possible states and starts in a start state, a set of accepting
+states, and a transition function that at each step allows tells the machine
+wether to write to the tape, move the head, or change states. There are
+modifications to this definition which allow for multiple tapes or
+nondeterminism, but they are all equivalent to the basic
+definition. Additionally, a \ac{PDA} with two stacks is equivalent to a Turing
+machine, as the stacks can be thought of as extending out in opposite
+directions, forming an infinite tape.
+
 \section{Undecidability}
+In the previous section we were concerned primarily with problems that our
+models of computation were able to solve. However, much as an \ac{FSA} is unable
+to solve the problem of whether a string is a palindrome, there are problems
+that are too difficult for even \aclp{TM} to solve. When a decision problem $A$
+cannot be solved by a model $\mathcal{M}$, we say that $A$ is undecidable for
+$\mathcal{M}$. When we just say that $A$ is undecidable, we mean that it is
+undecidable for \acp{TM}. \par
+It can be difficult to come up with obvious examples of problems that are
+undecidable. After all, we usually talk about sets based on the patterns they
+exhibit, and those patterns are typically simple enough to be enumerated by a
+computer. A promising place to look for problems that a model cannot decide is
+problems about the model itself, and the most famous undecidable problem is of
+that type. The Halting Problem, introduced by Alan Turing in the same paper
+where he introduced the Turing machine, asks whether it is possible to contsruct
+a Turing machine that can determine whether a Turing machine run on a given
+input will ever halt and return an
+answer~\cite{turing37_comput_number_with_applic_to_entsc}. The proof is through
+a method called \emph{diagonalizdation}, due to its similarity to Cantor's
+diagonal argument.
 
@@
   \label{prob:halting}
-  \cite{turing37_comput_number_with_applic_to_entsc}
+  Let $M$ be a Turing Machine and $i$ be an input. If $M$ is run on $i$, will it
+  eventually halt?
+  \begin{proof}[Proof of Undecidability]
+    Suppose for the sake of contradiction that the halting problem is
+    decidable. Then, there exists some machine
+    $\mathcal{O}$ that can decide the halting problem. Then, we can construct a
+    Turing machine $H$ that simulates $\mathcal{O}$ on its input $(M,i)$. In the
+    case where $\mathcal{O}(M,i)$ accepts, $H$ enters into an infinite
+    loop, whereas in the case where $\mathcal{O}(M,i)$ does not accept, $H$
+    halts. Consider $H(H,\epsilon)$. If $H$ halts, then
+    $\mathcal{O}(H,\epsilon)$ accepts, and so $H$ enters an infinite loop, and
+    so it does not halt. But if $H$ does not halt, then
+    $\mathcal{O}(H,\epsilon)$ does not accept, so $H$ halts. This is a
+    contradiction, so our premise that the halting problem is decidable must be false.
+  \end{proof}
+
   \cite{sipser13:_introd_theor_comput}
@@
 
-\begin{defn}
-  A Pushdown Automata is...
-\end{defn}
+% \begin{defn}
+%   A Pushdown Automata is...
+% \end{defn}
 
@@
 
-\subsection{Properties and Structure}
-\begin{lem}
-  The set $\mathcal{D}$ is of cardinality $2^{\aleph_0}$.
-\end{lem}
-
-\begin{lem}
-  Every Turing degree contains countably infinite elements.
-\end{lem}
-
 While it is simple to explicitly state problems lying in the lower degrees,
@@
 
+One of the earliest questions about the structure of the Turing degrees was
+Post's Problem, first posed in 1944~\cite{post44:_recur}. It was solved
+independently by Friedberg and Mucnik in the late 1950s using a technique
+called the ``priority method''~\cite{Friedberg236, muchnik1956unsolvability}. A solution that does not
+involve a priority argument was provided by Ku\v{c}era in 1986~\cite{Kucera:1986:APS:22416.22462}.
+
 \begin{prob}[Post's Problem]  
-  \cite{post44:_recur}
+  Does there exist a computably enumerable degree strictly between $0$ and $0'$?
   % Marchenkov 1976
@@
   % ``Natural'' solns don't exist
-  \cite{Friedberg236}
 \end{prob}
+
+\subsection{Properties and Structure}
+\begin{lem}
+  The set $\mathcal{D}$ is of cardinality $2^{\aleph_0}$.
+\end{lem}
+
+\begin{lem}
+  Every Turing degree contains countably infinite elements.
+\end{lem}
 

### 2018-08-14 paper.tex
jump operator definition

@@
 \usepackage{tikz}
-\usetikzlibrary{automata,positioning}
+\usetikzlibrary{automata,positioning,shapes}
 \usepackage[nolist]{acronym}
@@
   \acro{CFL}[CFL]{Context-Free Language}
+  \acro{DCFL}[DCFL]{Deterministic Context-Free Language}
 \end{acronym}
@@
 \cite{lerman16:_degrees_unsol}
+\todo{}
 \subsection{Decision Problems}
@@
   \end{proof}
-
-  \cite{sipser13:_introd_theor_comput}
 \end{prob}
-\begin{thm}[Rice's Theorem] \cite{kozen99_autom}
-  % Proof by reduction to the halting problem?
-\end{thm}
-\subsection{Incompleteness}
-\begin{thm}[G\"odel's Incompleteness Theorem] 
-  % Kleene 1943? Also shown in hopcroft 1979
-  \cite{kleene43_recur_predic_quant}
-\end{thm}
+% \begin{thm}[Rice's Theorem] \cite{kozen99_autom}
+%   % Proof by reduction to the halting problem?
+% \end{thm}
+% \subsection{Incompleteness}
+% \begin{thm}[G\"odel's Incompleteness Theorem] 
+%   % Kleene 1943? Also shown in hopcroft 1979
+%   \cite{kleene43_recur_predic_quant}
+% \end{thm}
 \subsection{Reducibility}
@@
 
-\section{Context Free Grammars}
+\section{Context Free Languages}
 
 \begin{defn}
- A Context Free Grammar is...
+ A \ac{CFG} is a finite set of nonterminal symbols, a finite set of terminal
+ symbols, a finite number of production rules that produce a string of terminal
+ and nonterminal symbols from a nonterminal symbol, and a starting symbol.
 \end{defn}
 
-% \begin{defn}
-%   A Pushdown Automata is...
-% \end{defn}
+\begin{defn}
+  The \emph{language} of a \ac{CFG} is the set of words that can be formed by
+  the successive application of production rules to the start symbol. These
+  languages are called \acp{CFL}.
+\end{defn}
 
-\begin{thm}[Equivalence of CFGs and Pushdown Automata]
-  \cite{hopcroft07:_introd_autom_theor_languag_comput}
+\acp{CFG} provide another useful model of computation. However, the languages
+that they are able to
+express are exactly the same set of languages as \acp{PDA}~\cite{hopcroft07:_introd_autom_theor_languag_comput}. This equivalence allows
+us to choose between the two models and use the one that is most useful for
+proving facts about these languages.
+
+\begin{thm}[Closure Properties of \acp{CFL}]
+  \aclp{CFG} have a variety of closure properties:
+  \begin{itemize}
+  \item Let $L_1,L_2$ be \acp{CFL}. Then $L_1\cup L_2$ is a \ac{CFL}.
+    \begin{proof}
+      The languages $L_1$ and $L_2$ are \aclp{CFL}, so there exist \aclp{PDA} $P_1$ and $P_2$
+      that decide them. Consider the \ac{PDA} $U$ shown in
+      Figure~\ref{pda:union}, which simulates $P_1$ and $P_2$ on the input. $U$
+      is a \ac{PDA} that decides $L_1\cup L_2$, so it follows that
+      $L_1\cup L_2$ is a \ac{CFL}.
+      \begin{figure}[h]
+        \caption{$U$, a \ac{PDA} that simulates $P_1$ and $P_2$.}
+        \label{pda:union}
+         \begin{tikzpicture}[shorten >=1pt,node distance=3cm,on grid,auto]
+  \node[state,initial] (q_0) {$q_0$};
+  \node[state,diamond] (P_1) [above right=of q_0]{$P_1$};
+  \node[state,diamond] (P_2) [below right=of q_0]{$P_2$};
+  \draw[->]
+  (q_0) edge node {$\epsilon,\epsilon\to\epsilon$} (P_1);
+  \draw[->]
+  (q_0)
+  edge node {$\epsilon,\epsilon\to\epsilon$} (P_2);
+  \end{tikzpicture}
+      \end{figure}
+    \end{proof}
+  \item Let $L_1,L_2$ be \acp{CFL}. Then the concatenation \[
+      \{w_1w_2 \mid w_1\in L_1, w_2 \in L_2\}
+    \]
+    is a \ac{CFL}.
+    \begin{proof}
+      The languages $L_1$ and $L_2$ are \aclp{CFL}, so there exist \acp{CFG}
+      $G_1,G_2$ with start symbols $S_1,S_2$ that produce them. Suppose, without loss of generality, that
+      $G_1,G_2$ have no nonterminal symbols in common (if they do, then we can rename
+      them). Consider the grammar $S \rightarrow S_1S_2$. Then, this is a
+      context free grammar, and its language is the concatenation of $L_1$ and
+      $L_2$. So, the concatenation of $L_1$ and $L_2$ is a \ac{CFL}.
+    \end{proof}
+  \item Let $L$ be a \ac{CFL}. Then,
+    \[
+      L^* = \bigcup_{i \geq 0}\{w_1 \dots w_k\dots w_i \mid w_k\in L\}
+    \]
+    is a \ac{CFL}.
+    \begin{proof}
+      The language $L$ is a \ac{CFL}, so there exists a \ac{CFG} $G$ with start
+      symbol $S_1$ that produces it. Consider the grammar given by $S\rightarrow
+      S_1S\mid\epsilon$. This is a \ac{CFG} that produces $L^*$, so it follows
+      that $L^*$ is a \ac{CFL}.
+    \end{proof}
+    \item Intersection with regular languages~\cite{hopcroft07:_introd_autom_theor_languag_comput}.\todo{}
+  \end{itemize}
 \end{thm}
-\begin{thm}[Closure Properties]
-  \cite{sipser13:_introd_theor_comput}
-  Union, Concat, Star, intersection with regular, substitution
-\end{thm}
+
 \begin{prob}[Emptiness]
@@
   Let $G$ be a context-free grammar. Is $L(G)=\varnothing$?
+  \begin{proof}[Proof of Decidability]
+    \todo{}
+  \end{proof}
 \end{prob}
@@
   Let $G$ be a context-free grammar. Is $L(G)$ finite?
+  \begin{proof}[Proof of Decidability]
+    \todo{}
+  \end{proof}
 \end{prob}
@@
 \begin{defn}
-An ambiguous grammar is...
+We say that a \acl{CFG} is \emph{ambiguous} if there is more than one way to
+apply the production rules to reach an accepted word, assuming a production rule
+is always applied to the leftmost nonterminal symbol. This is equivalent to
+stating that there is more than one parse tree that parses an accepted word.
 \end{defn}
@@
 
-\subsubsection{Problems decidable for Deterministic CFLs}
+\subsubsection{Problems decidable for DCFLs}
 \cite{ginsburg65:_deter}
 \begin{defn}
-  A deterministic context-free language is a language accepted by a
-  deterministic pushdown automata.
+  A \ac{DCFL} is a language accepted by a
+  \acl{DPDA}.
 \end{defn}
@@
 \begin{thm}[Closure Properties]
-  If $L$ is a deterministic context-free language, then $\overline{L}$ is a
-  deterministic context-free language.
+  \label{dcfl:complement}
+  If $L$ is \ac{DCFL}, then $\overline{L}$ is a
+  \ac{DCFL}. \todo{}
 \end{thm}
@@
 \begin{prob}
-  Let $L$ be a deterministic context-free language. Is $L = \Sigma^*$?
+  Let $L$ be a \ac{DCFL}. Is $L = \Sigma^*$?
+  \begin{proof}[Proof of Decidability]
+    We have that $L$ is a \ac{DCFL}, so by Theorem~\ref{dcfl:complement},
+    $\overline{L}$ is also a $\ac{DCFL}$. By Problem~\ref{prob:cfg:emptiness},
+    we have that whether $\overline{L}$ is empty is decidable, so it follows
+    that the question of whether $L = \Sigma^*$ is also decidable.
+  \end{proof}
 \end{prob}
@@
   Let $L_1,L_2$ be deterministic context-free languages. Is $L_1=L_2$?
-
+\todo{}
 \end{prob}
@@
 \begin{defn}
-  Computably Enumerable
+  An oracle machine for a degree $A$, denoted $\mathcal{O}_A$, is a Turing
+  machine equipped with a function that is able to decide $A$.
 \end{defn}
@@
 \begin{defn}
-  Completeness
+  A problem $P$ is called \emph{computably enumerable} or \emph{semidecidable}
+  for some degree $d\in\mathcal{D}$ if there exists a problem $A\in d$ such that
+  it is possible to construct an oracle machine $\mathcal{O}_A$ such that
+  $\mathcal{O}_A$ halts for all the inputs where $P$ is true, and does not halt
+  on all the inputs for which $P$ is false. If $A=\varnothing$, and so the
+  oracle machine does not have an oracle, then we simply say that $P$ is
+  computably enumerable, semidecidable, or \emph{Turing recognizable}.
 \end{defn}
+\begin{rem}
+  The Halting problem, Problem~\ref{prob:halting}, is Turing recognizable. The
+  machine that recognizes the problem simply simulates the input machine.
+\end{rem}
 
@@
 \begin{defn}
-  The jump operator
+  The jump of $A$, denoted $A'$, is degree of the halting problem for oracle
+  machines for $A$.
 \end{defn}
+
+The jump operator has many properties that make it convenient for studying the
+Turing degrees, which are listed below. Further commentary on the jump operator
+and its properties is available in Robert Soare's book \emph{History of Computability}~\cite{soare16_turin_comput}.
+\begin{fact}
+  The jump operator has many properties that make it convenient for studying the
+Turing degrees, some of which are listed below. Further commentary on the jump operator
+and its properties is available in Robert Soare's book \emph{History of Computability}~\cite{soare16_turin_comput}.
+  \begin{itemize}
+  \item $A'$ is computably enumerable in $A$.
+  \item $A' \not\leq_T A$.
+  \item If $B=_TA$ then $B'=_TA'$.
+  \end{itemize}
+\end{fact}
 
@@
 % Myhill 56 c.e.lattice
+\begin{defn}
+  A partially ordered set $S$ is said to form an \emph{upper semilattice} if, for every
+ $x,y\in S$, there exists some element $z\in S$ such that $z \geq x,y$ and
+ for all $w\in A$ such that $w\geq x,y$, we have that $w \geq z$. In other
+ words, $x$ and $y$ have a supremum.
+\end{defn}
 
-\begin{lem}
-  D does not form a lattice
-  \cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
-\end{lem}
+\begin{defn}
+  A partially ordered set $S$ is said to form a \emph{lattice} if all pairs
+  $\{x,y\}\in S$ have both an infinum and a supremum.
+\end{defn}
+
 \begin{thm}
-  D forms an upper semi-lattice
+  $\mathcal{D}$ forms an upper semilattice.
   % Kleene-Post 54 Degrees upper semi-lattice, jump an operator on degrees
@@
 \end{thm}
+\begin{lem}
+ $ \mathcal{D}$ does not form a lattice.
+  \cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
+\end{lem}
 
@@
 
-\subsection{Turing Degrees of Problems related to CFGs}
-\cite{REEDY197577}
+% \subsection{Turing Degrees of Problems related to CFGs}
+% \cite{REEDY197577}
 

### 2018-08-15 paper.tex
Degrees of cardinality continuum

@@
 \subsection{Decision Problems}
+What does it mean to \emph{solve a problem}? What is a \emph{problem}, anyways?
+These questions may seem superfluous, they must be addressed in order to build a
+well defined theory of the difficulty of problems. In this paper, we discuss a
+specific type of problem, the \emph{decision problem}. Decision problems are
+essentially sets of natural numbers, and to solve the problem is to be able to
+decide whether any given element is inside the set or not. Because set theory is
+commonly used as the foundational system of mathematics, we can convert problems
+expressed in terms of more complex mathematical objects into decision problems,
+and thus by studying decision problems can build a system that is able to assess
+the computability of all mathematical problems. For example, we can express
+strings over the alphabet $\{0,1\}$ as sets of natural numbers, and thus can
+study problems about languages of such strings with the same tools we use for
+decision problems. Going further, we can express the source code of a computer
+program as a string over the alphabet $\{0,1\}$, and so can consider questions
+about the behavior of computer programs.
 
@@
   If $L$ is \ac{DCFL}, then $\overline{L}$ is a
-  \ac{DCFL}. \todo{}
+  \ac{DCFL}.
+  \begin{proof}
+    Let $L$ be a \ac{DCFL}. Then, there exists some \ac{DPDA} $P$ that
+    recognizes $L$. We can construct a new \ac{DPDA} $P'$ by complementing the
+    accepting states of $P$. Because $P$ is a \ac{DPDA} and thus is only in one
+    state at a time, $P'$ will recognize $\overline{L}$ and thus $\overline{L}$
+    is a \ac{DCFL}. This would not hold if $P$ were only a \ac{PDA}, since a
+    nondeterministic \ac{PDA} accepts if \emph{any} branch ends on an accepting
+    state, and thus complementing the accepting states would not recognize the
+    complement of the originally recognized language.
+\end{proof}
 \end{thm}
@@
 
-\begin{lem}
+\begin{fact}
   The relation $\leq_T$ is a partial ordering of $\mathcal{D}$.
-\end{lem}
+\end{fact}
 
@@
 \subsection{Properties and Structure}
+
 \begin{lem}
-  The set $\mathcal{D}$ is of cardinality $2^{\aleph_0}$.
+  \label{lem:deg_count}
+  Every Turing degree is of cardinality $\aleph_0$.
 \end{lem}
@@
 \begin{lem}
-  Every Turing degree contains countably infinite elements.
+  The set $\mathcal{D}$ is of cardinality $2^{\aleph_0}$.
+  \begin{proof}
+    First, observe that the cardinality of $\mathcal{D}$ is at most
+    $2^{\aleph_0}$, since it is a set of sets of decision problems and there are
+    at most $2^{\aleph_0}$ decision problems since decision problems are subsets
+    of $\mathbb{N}$.\par Now, recall from Lemma~\ref{lem:deg_count} that each
+    degree contains $\aleph_0$ decision problems. From this, it follows that
+    there must be at least $2^{\aleph_0}$ degrees. So, we have that $\lvert \mathcal{D} \rvert=2^{\aleph_0}$.
+  \end{proof}
 \end{lem}

### 2018-08-15 paper.tex
Upper Semilattice result

@@
   Every Turing degree is of cardinality $\aleph_0$.
+  \begin{proof}
+    We have that since every Turing machine can be expressed as a finite string over
+    $\{0,1\}$, it can be expressed as a finite subset of $\mathbb{N}$. So, it follows
+    that for any given oracle, there are at most $\aleph_0$ oracle machines, so
+    at most $\aleph_0$ problems are decidable for any Turing degree.
+    \par
+    Now that an upper bound has been established,  we show that the cardinality
+    of each degree is exactly $\aleph_0$.\todo{}
+  \end{proof}
 \end{lem}
@@
 \begin{thm}
+  \label{thm:up_sem}
   $\mathcal{D}$ forms an upper semilattice.
   % Kleene-Post 54 Degrees upper semi-lattice, jump an operator on degrees
-  \cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
+    \begin{proof}
+    Observe that it is sufficient to prove that every two degrees have a
+    supremum. Let $A,B\in \mathcal{D}$ and  $P_1\in A, P_2 \in B$. Consider the operation
+    \[
+      A\vee B = \deg\left( \{2n \mid n \in P_1\}\cup\{2n+1\cup n\in P_2\} \right),
+    \]
+    where $\deg(A)$ is the Turing degree of $A$. Now, we show that $A\vee B = \sup(A,B)$.\par First, we show that
+    $A,B \leq_T A \vee B$. This is equivalent to showing
+    that problems in $A$ and $B$ are decidable by oracle machines for $A\vee
+    B$. But simple construction gives oracle machines that can decide $P_1,P_2$
+    given an oracle for $A\vee B$: for any $n$, if $2n\in A\vee B$ then
+    $n\in P_1$, and if $2n+1\in A\vee B$ then $n\in P_2$. This holds for any
+    $P_1,P_2$, so it follows that $A,B \leq_T A \vee
+    B$. \par Now, let $W\in \mathcal{D}$ such that $W \geq
+    A,B$ and let $P\in A\vee B$. Then, we have that for every $n$, we can
+    determine whether $n\in P$ by determining whether $n$ is even and
+    $\frac{n}{2}\inP_1$ or odd and $\frac{n-1}{2}\in P_2$. Since both the parity
+    of a number and $P_1$ and $P_2$ are decidable in $W$, it follows that
+    $A\veeB$ is decidable in $W$, and thus $A\vee B \leq_T W$. So, it follows
+    that $A\vee B = \sup(A,B)$. This holds for all $A,B\in\mathcal{D}$, so it
+    follows that $\mathcal{D}$ forms an upper semilattice.  
+  \end{proof}
 \end{thm}
+
 \begin{lem}
+  \label{lem:not_lat}
   $ \mathcal{D}$ does not form a lattice.
-  \cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
+
 \end{lem}
+\begin{rem}
+Theorem~\ref{thm:up_sem} and Lemma~\ref{lem:not_lat} were first shown in
+Kleene's \emph{The Upper Semi-Lattice of Degrees of Recursive Unsolvability}~\cite{kleene54_upper_semi_lattic_degrees_recur_unsol}.
+\end{rem}
 

### 2018-08-15 paper.tex
Citing not a lattice result

@@
     determine whether $n\in P$ by determining whether $n$ is even and
-    $\frac{n}{2}\inP_1$ or odd and $\frac{n-1}{2}\in P_2$. Since both the parity
+    $\frac{n}{2}\in P_1$ or odd and $\frac{n-1}{2}\in P_2$. Since both the parity
     of a number and $P_1$ and $P_2$ are decidable in $W$, it follows that
-    $A\veeB$ is decidable in $W$, and thus $A\vee B \leq_T W$. So, it follows
+    $A\vee B$ is decidable in $W$, and thus $A\vee B \leq_T W$. So, it follows
     that $A\vee B = \sup(A,B)$. This holds for all $A,B\in\mathcal{D}$, so it
@@
 
-\begin{lem}
-  \label{lem:not_lat}
+\begin{fact}
+  \label{fact:not_lat}
   $ \mathcal{D}$ does not form a lattice.
-
-\end{lem}
+\end{fact}
 \begin{rem}
-Theorem~\ref{thm:up_sem} and Lemma~\ref{lem:not_lat} were first shown in
-Kleene's \emph{The Upper Semi-Lattice of Degrees of Recursive Unsolvability}~\cite{kleene54_upper_semi_lattic_degrees_recur_unsol}.
+Theorem~\ref{thm:up_sem} and Fact~\ref{fact:not_lat} were first shown in
+Kleene's \emph{The Upper Semi-Lattice of Degrees of Recursive
+  Unsolvability}~\cite{kleene54_upper_semi_lattic_degrees_recur_unsol}. A simpler proof
+of Fact~\ref{fact:not_lat} is available in Robert Soare's book \emph{Turing Computability}~\cite[pp.~142--144]{soare16_turin_comput}.
 \end{rem}

### 2018-08-15 paper.tex
further reading, minimal degrees, removed roadmap

@@
 These questions may seem superfluous, they must be addressed in order to build a
-well defined theory of the difficulty of problems. In this paper, we discuss a
+well defined theory of the difficulty and computability of problems. In this paper, we discuss a
 specific type of problem, the \emph{decision problem}. Decision problems are
@@
 
-The jump operator has many properties that make it convenient for studying the
-Turing degrees, which are listed below. Further commentary on the jump operator
-and its properties is available in Robert Soare's book \emph{History of Computability}~\cite{soare16_turin_comput}.
 \begin{fact}
@@
 Turing degrees, some of which are listed below. Further commentary on the jump operator
-and its properties is available in Robert Soare's book \emph{History of Computability}~\cite{soare16_turin_comput}.
+and its properties is available in Robert Soare's book \emph{Turing
+  Computability}~\cite{soare16_turin_comput}, as well as in Shore and Slaman's
+paper \emph{Defining the Turing Jump}~\cite{shore99:_defin_turin_jump}.
   \begin{itemize}
@@
 \end{rem}
+ A final result respecting the structure of $\mathcal{D}$ is the existence of
+ \emph{minimal degrees}. A degree $A$ is minimal if $A \neq_T 0$ ($A$ is not
+ Turing-decidable), but there does not exist any degree $B\in \mathcal{D}$ such
+ that $0 <_T B <_T A$. The existence of such degrees was first shown by Spector
+ in 1956~\cite{spector56_degrees_recur_unsol}. An alternative proof is by
+ Shoenfield in 1966~\cite{shoenfield66_theor_minim_degrees}.
 
-\begin{thm}
-  Existence of minimal degrees
-  % Spector 56,Shoenfield 66 (Better)
-  \cite{spector56_degrees_recur_unsol}
-  \cite{shoenfield66_theor_minim_degrees}
-\end{thm}
+\section{Further Reading}
+The study of the Turing degrees was a very active topic in the late
+20\textsuperscript{th} and early 21\textsuperscript{st} centuries, and this
+paper has just scratched the surface of the field. \par
+Readers particularly interested in the structure of the computably enumerable
+degrees might be interested in a result by Sacks in 1964 showing their
+density~\cite{sacks64:_recur_enumer_degrees_dense}, while those interested in
+\aclp{CFG} or the Turing degrees of problems about them should consult \cite{REEDY197577} and
+\cite{ginsburg66:_mathem_theor_contex_languag}
+\cite{hopcroft07:_introd_autom_theor_languag_comput}\todo{chapter for cfgs}. For excellent, comprehensive,
+and more rigorous textbooks on the study of Turing computability and the Turing
+degrees, consider \cite{soare16_turin_comput} \cite{lerman16:_degrees_unsol} or
+the classic text \cite{kleene80_introd}.
 
-\begin{thm}
-  The c.e. degrees are dense
-  \cite{sacks64:_recur_enumer_degrees_dense}
-\end{thm}
+% \begin{thm}
+%   The c.e. degrees are dense
+%   \cite{sacks64:_recur_enumer_degrees_dense}
+% \end{thm}
 
-Homogeneity problems
-% Feiner 1970, Shore 1979
+% Homogeneity problems
+% % Feiner 1970, Shore 1979
 
-Theory of D Undecidable
+% Theory of D Undecidable
 
-Theory of D equivalent to to second-order arithmetic \cite{simpson77:_first_order_theor_degrees_recur_unsol}
+% Theory of D equivalent to to second-order arithmetic \cite{simpson77:_first_order_theor_degrees_recur_unsol}
 

### 2018-08-16 paper.tex
Cleaning up Senizergues, further reading sections

@@
 
-The decidability of the following problem was an open problem in the field of
-computability theory from 1965, when it was introduced by Ginsburg and Greibach \cite{ginsburg65:_deter}. In 1997, when it was solved by G\'eraud S\'enizergues
-\cite{senizergues_det_pd_decid}. A sketch of the proof follows.
-
-\begin{prob}[The Equivalence Problem for Deterministic CFGs]
-  Let $L_1,L_2$ be deterministic context-free languages. Is $L_1=L_2$?
-\todo{}
-\end{prob}
+Finally, the decidability of the equality problem for \acp{DCFL} was an open problem in the field of
+computability theory from 1965, when it was introduced by Ginsburg and Greibach
+until 1997, when it was shown to be decidable by G\'eraud S\'enizergues
+\cite{ginsburg65:_deter,senizergues_det_pd_decid}. S\'enizergues received the 2002 G\"odel Prize
+for the result.
 
@@
 density~\cite{sacks64:_recur_enumer_degrees_dense}, while those interested in
-\aclp{CFG} or the Turing degrees of problems about them should consult \cite{REEDY197577} and
-\cite{ginsburg66:_mathem_theor_contex_languag}
-\cite{hopcroft07:_introd_autom_theor_languag_comput}\todo{chapter for cfgs}. For excellent, comprehensive,
+\aclp{CFG} or the Turing degrees of problems about them should consult Reedy's
+1975 paper on the Turing degree of a specific problem~\cite{REEDY197577} and
+Ginsburg's book \emph{The Mathematical Theory of Context Free
+  Languages}~\cite{ginsburg66:_mathem_theor_contex_languag}. Additionally, a
+simple overview of the field is available in Hopcroft's classic text on
+computability
+theory~\cite[chs.~5--7]{hopcroft07:_introd_autom_theor_languag_comput}. \par For excellent, comprehensive,
 and more rigorous textbooks on the study of Turing computability and the Turing
-degrees, consider \cite{soare16_turin_comput} \cite{lerman16:_degrees_unsol} or
-the classic text \cite{kleene80_introd}.
+degrees, consider Soare and Lerman's
+books~\cite{soare16_turin_comput,lerman16:_degrees_unsol}, as well as
+Kleene's original text on the subject~\cite{kleene80_introd}.
 

### 2018-08-17 paper.tex
First complete draft!

@@
 \section{Introduction}
-Historical overview of the field, motivation for study
-\cite{ambos-spies06:_degrees_unsol}
-\cite{soare16_turin_comput}
-\cite{soare1999history}
-\cite{lerman16:_degrees_unsol}
-\todo{}
+
+The field of computability theory focuses on abstract models of computation and
+the sorts of problems that they are able to solve. Unlike complexity theory,
+which is focused on the resources required to perform a computation,
+computability theory focuses only on whether a certain computation is possible
+on a given machine. The practical application of the theory focuses on models of
+computation of equivalent power to general purpose computers, allowing for it to
+be determined whether it possible to solve specific problems on actual
+machines. The Church-Turing thesis, a significant conjecture in the field,
+states that the model that we use to understand the power of general purpose
+computers has the same power as humans applying algorithms. If true, it means
+that it is not possible to design a computational process that is able to solve
+more problems than a general-purpose computer. However, even if it is true, it is still possible to
+imagine models that have more power than general purpose computers by endowing
+them with \emph{oracles} the ability to evaluate problems that we know
+general-purpose computers cannot. Because of this, computability is still able
+to draw distinctions in the difficulty of problems too difficult for
+general-purpose computers to solve, dividing them into \emph{Turing degrees}, or
+\emph{degrees of unsolvability}. Additionally, it is possible to construct
+models of computation with less power than general-purpose computers, and thus
+classify the difficulty of problems that general-purpose computers can solve. We
+can also show the differences in difficulty of problems simply by showing that
+it is possible to use the answer of one problem to solve another, a technique
+called \emph{reduction}. So, it is possible to create an infinite hierarchy
+describing the difficulty of problems, and to investigate the sort of structure
+this hierarchy takes.
+
+% Historical overview of the field, motivation for study
+
+% \cite{soare16_turin_comput}
+% \cite{soare1999history}
+% \cite{lerman16:_degrees_unsol}
+% \todo{}
 \subsection{Decision Problems}
 What does it mean to \emph{solve a problem}? What is a \emph{problem}, anyways?
-These questions may seem superfluous, they must be addressed in order to build a
+These questions may seem esoteric, they must be addressed in order to build a
 well defined theory of the difficulty and computability of problems. In this paper, we discuss a
@@
 decidable, it is sufficient to show that it is reducible to some problem $B$
-that is known to be decidable.
-\cite{sipser13:_introd_theor_comput}
-\cite{post44:_recur}
-
-\cite{kleene80_introd}
+that is known to be decidable~\cite{sipser13:_introd_theor_comput,post44:_recur,kleene80_introd}.
 
@@
     \end{proof}
-    \item Intersection with regular languages~\cite{hopcroft07:_introd_autom_theor_languag_comput}.\todo{}
+    \item Intersection with regular languages~\cite{hopcroft07:_introd_autom_theor_languag_comput}.
   \end{itemize}
@@
 
-\begin{prob}[Emptiness]
-  \label{prob:cfg:emptiness}
-  Let $G$ be a context-free grammar. Is $L(G)=\varnothing$?
-  \begin{proof}[Proof of Decidability]
-    \todo{}
-  \end{proof}
-\end{prob}
-
-\begin{prob}[Finite]
-  Let $G$ be a context-free grammar. Is $L(G)$ finite?
-  \begin{proof}[Proof of Decidability]
-    \todo{}
-  \end{proof}
-\end{prob}
+We have that the emptiness and finiteness problems are decidable for
+\acp{CFG}~\cite{hopcroft07:_introd_autom_theor_languag_comput,
+  sipser13:_introd_theor_comput}. Another problem that is decidable for
+\acp{CFG} is the question of whether a given \ac{CFL} is wholly contained by a
+given regular language.
 
@@
   language. So the problem of containment by a regular language is reducible to
-  Problem~\ref{prob:cfg:emptiness}, the problem of emptiness, which is
+  the problem of emptiness, which we know to be
   decidable. So, the problem of containment by a regular language is decidable.
@@
 \begin{prob}[The Post Correspondence Problem]
+  \label{prob:pcp}
   Consider some alphabet $\Sigma$ and two finite lists of words over $\Sigma$
@@
 \begin{prob}[Disjointness]
+  \label{prob:disjointness}
   Let $P,Q$ be context-free grammars. Is $L(P)\cap L(Q) = \varnothing$?
@@
   \]
-  \begin{proof}[Proof of Undecidability]\todo{}
-
+  \begin{proof}[Proof of Undecidability]
+    Consider the Post correspondence problem for two lists of words
+    $A,B$. Construct their corresponding grammars $G_A,G_B$ as in Problem~\ref{prob:disjointness}. Then, the languages
+    of these grammars are not just \acp{CFL}, but are also \aclp{DCFL},
+    discussed in detail in Subsection~\ref{sub:dcfl}, since we
+    can construct \aclp{DPDA} to accept them. We have that if
+    \[\overline{L(G_A)\cap L(G_B)} = \Sigma^*,\] then
+    \[L(G_A)\cap L(G_B)=\varnothing.\] Since these are grammars corresponding to
+    lists in a Post correspondence problem, it follows
+    that \[\overline{L(G_A)\cap L(G_B)} = \Sigma^*\] implies that there is no
+    solution to that Post correspondence problem. Additionally, by DeMorgan's Law, we have
+    that
+    \[
+      \overline{L(G_A)\cap L(G_B)} =
+      \overline{L(G_A)}\cup\overline{L(G_B)}.
+    \]
+    Since \acp{DCFL} are closed under complement and \acp{CFL} are closed under
+    union, it follows that $ \overline{L(G_A)\cap L(G_B)}$ is a \ac{CFL}, so the
+    question of whether it is empty is decidable. So, it follows that the
+    problem of universality reduces to Problem~\ref{prob:pcp}, the Post
+    correspondence problem, which is undecidable. So, we have that the
+    problem of universality is undecidable.
   \end{proof}
@@
     Consider the Post correspondence problem for two lists of words
-    $A,B$. Construct their corresponding grammars $G_A,G_B$. Now, consider the
+    $A,B$. Construct their corresponding grammars $G_A,G_B$ as in Problem~\ref{prob:disjointness}. Now, consider the
     grammar
@@
 \end{prob}
-\cite{greibach66:_unsol_recog_linear_contex_free_languag}
-\cite{Hopcroft1969}
+
 \begin{prob}
@@
 
-\subsubsection{Problems decidable for DCFLs}
-\cite{ginsburg65:_deter}
+\subsection{Problems decidable for DCFLs}
+\label{sub:dcfl}
 \begin{defn}
@@
 \end{defn}
+\acp{DCFL} are another important classification of languages, and are a proper
+subset of \acp{CFL}. They are discussed in detail by Ginsburg~\cite{ginsburg65:_deter}.
 
@@
     We have that $L$ is a \ac{DCFL}, so by Theorem~\ref{dcfl:complement},
-    $\overline{L}$ is also a $\ac{DCFL}$. By Problem~\ref{prob:cfg:emptiness},
-    we have that whether $\overline{L}$ is empty is decidable, so it follows
+    $\overline{L}$ is also a $\ac{DCFL}$.
+    We know that whether $\overline{L}$ is empty is decidable, so it follows
     that the question of whether $L = \Sigma^*$ is also decidable.
@@
 One of the earliest questions about the structure of the Turing degrees was
-Post's Problem, first posed in 1944~\cite{post44:_recur}. It was solved
+Post's Problem, first posed in 1944~\cite{post44:_recur}. Post's Problem asks
+whether there is a computably enumerable degree strictly between $0$ and $0'$. It was solved
 independently by Friedberg and Mucnik in the late 1950s using a technique
@@
 
-\begin{prob}[Post's Problem]  
-  Does there exist a computably enumerable degree strictly between $0$ and $0'$?
-  % Marchenkov 1976
-  % Independently by Friedberg 57, Mucnik 56/58 using priority method
-  % Priority free solution Kucer 1986''
-  % ``Natural'' solns don't exist
-\end{prob}
+% \begin{prob}[Post's Problem]  
+%   Does there exist a computably enumerable degree strictly between $0$ and $0'$?
+%   % Marchenkov 1976
+%   % Independently by Friedberg 57, Mucnik 56/58 using priority method
+%   % Priority free solution Kucer 1986''
+%   % ``Natural'' solns don't exist
+% \end{prob}
 
@@
     Now that an upper bound has been established, we show that the cardinality
-    of each degree is exactly $\aleph_0$.\todo{}
+    of each degree is exactly $\aleph_0$. Suppose, for the sake of contradiction,
+    that a Turing degree contained only a finite number of decision
+    problems. Then, the degree would be decidable, since every finite set is
+    decidable. But the only decidable degree is $0$, and in the previous section we
+    showed that the cardinality of $0$ is $\aleph_0$, so it follows that every
+    degree has cardinality $\aleph_0$.
   \end{proof}
@@
 books~\cite{soare16_turin_comput,lerman16:_degrees_unsol}, as well as
-Kleene's original text on the subject~\cite{kleene80_introd}.
+Kleene's original text on the subject~\cite{kleene80_introd}. \par
+Finally, Klaus Ambos-Spies and Peter Fejer's unpublished \emph{Degrees
+  of Unsolvability} provides an exhaustive overview of the history of the study
+of the Turing degrees~\cite{ambos-spies06:_degrees_unsol}.
 

### 2018-08-23 paper.tex
fixing typo
Implementing Ronno's suggestions

@@
 \subsection{Decision Problems}
-What does it mean to \emph{solve a problem}? What is a \emph{problem}, anyways?
+What does it mean to \emph{solve a problem}? What is a \emph{problem}?
 These questions may seem esoteric, they must be addressed in order to build a
@@
 about the behavior of computer programs.
+
+\subsection{Overview}
+This paper aims to develop the ideas presented in the introduction in a natural
+manner, and at a level of rigor accessible to an interested undergraduate. We
+begin by presenting a few simple models of computation and assessing their
+ability to compute various sets. We then introduce the Turing Machine, and
+mention alternative models of equivalent power. Next, we introduce the concepts
+of undecidability and Turing reducibility, and present the halting problem as an
+example of an undecidable problem. In the interest of developing a strong
+intuition regarding the concept of undecidability, we examine a variety of
+decidable and undecidable problems about \aclp{CFG}. Finally, we employ the
+Turing reduction to introduce a structure on the set of decision problems, and
+present some basic results about its properties.
 
@@
 \ac{FSA} accepts the input if the sequential application of each element to the
-input finishes in an accepting state, and rejects otherwise. \aclp{FSA} are often represented by charts as in Figure~\ref{fig:fsa}, a
-representation of an \ac{FSA} that decides the first problem.
+input finishes in an accepting state, and rejects otherwise. \aclp{FSA} are often represented by diagrams as in Figure~\ref{fig:fsa}, a
+representation of an \ac{FSA} that decides the first problem. An explanation of
+the conventions used in these diagrams is available in Sipser's
+\emph{Introduction to the Theory of Computation}~\cite[][pp. 34,114]{sipser13:_introd_theor_comput}
 \begin{figure}[h]
@@
 that decides palindromes. In this representation, $\epsilon$ is an empty input,
-and  is a special symbol that denotes the bottom of the stack. The moves are
+and \$ is a special symbol that denotes the bottom of the stack. The moves are
 represented by a double of input and stack operation.
@@
   Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
-  to $B$ and write $A \leq_T B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
-  a machine $D_A$ that decides $A$.
+  to $B$ and write $A \leq_T B$ if it is possible to construct an oracle
+  machine for $B$ that decides $A$ (see
+  Definition~\ref{defn:oracle}).
 \end{defn}
@@
 \begin{defn}
+  \label{defn:oracle}
   An oracle machine for a degree $A$, denoted $\mathcal{O}_A$, is a Turing
@@
 Post's Problem, first posed in 1944~\cite{post44:_recur}. Post's Problem asks
-whether there is a computably enumerable degree strictly between $0$ and $0'$. It was solved
-independently by Friedberg and Mucnik in the late 1950s using a technique
-called the ``priority method''~\cite{Friedberg236, muchnik1956unsolvability}. A solution that does not
-involve a priority argument was provided by Ku\v{c}era in 1986~\cite{Kucera:1986:APS:22416.22462}.
+whether there is a computably enumerable degree strictly between $0$ and $0'$. A
+positive answer to the question was found independently by Friedberg and Mucnik in the late
+1950s using a technique called the ``priority method''~\cite{Friedberg236,
+  muchnik1956unsolvability}. A solution that does not involve a priority
+argument was provided by Ku\v{c}era in 1986~\cite{Kucera:1986:APS:22416.22462}.
 
@@
     problems. Then, the degree would be decidable, since every finite set is
-    decidable. But the only decidable degree is $0$, and in the previous section we
+    decidable. But we know that the only decidable degree is $0$, since all
+    decidable problems are mutually reducibile, and in the previous section we
     showed that the cardinality of $0$ is $\aleph_0$, so it follows that every
